Everything you wanted to know about PanDA but were truly truly afraid to ask

### 2017-03-05 PandaEvolution.tex
First pass IV (former IV and V) + restructuring

\subsection{Brief Introduction to Titan}

While PanDa  currently uses  more than 100,000  cores at well over 100 Grid
sites with a  peak performance  of 0.3 petaFLOPS,  next LHC data taking run will
require more resources  than Grid computing can possibly provide. The Worldwide
LHC Computing Grid (WLCG) infrastructure will be sufficient for the planned
analysis and data processing, but it will be insufficient  for Monte Carlo (MC)
production and any extra activities. Additional computing and storage  resources
are therefore required.  To alleviate  these challenges, ATLAS is engaged  in an
ambitious  program to expand  the current computing model to include  additional
resources such as the opportunistic   use of supercomputers   as well as
commercial and academic clouds. In turn this activity drives evolution of the
PanDA WMS\@.

\subsection{Use of Supercomputers with PanDA}

The idea is to extend the ATLAS Computing Model beyond the Grid into the domains
of supercomputers. Modern High Performance Computing  (HPC) involves  a very
large number of worker nodes  connected  through a  high-speed   network. The
worker nodes can have  multicore CPUs that can be augmented with massively
parallel Graphics Processing Units (GPUs) or other types of coprocessors. On an
HPC machine, a  typical job is highly parallel and connected,  where each core
calculates  a small part of the problem and coordinates the activity via message
passing interface (MPI). While this is different from the HEP computing
paradigm, where jobs are independent, it still shares common  features  such as
the use of parallelization. It is not a requirement  that HPC machines are able
to run any possible task, nor is it relevant how many kinds of job types that
can be run. What matters is the total number of cycles that can be offloaded
from the Grid. Standard  ATLAS  workflow  can not be easily ported to
supercomputers   due  to   several   complications such  as specialized worker
node  setups, no  outbound network connections,  limited memory per node, custom
operating systems,  etc. A reorganization  of the standard  workflow is
therefore needed.

The most suitable task for HPCs in HEP is event generation. Event generators are
mostly computational,  stand-alone code, with few input requirements.  There is
no need  to stage-in much data. The event generation  in ATLAS corresponds  to
10--15\% of all jobs on the Grid. A more difficult workload to adapt  for HPCs
is detector  simulation. Simulation tasks tend to be tied to the framework of a
particular simulation and often require  database access for geometry and
relevant run conditions.  However,  there is a  lot to be gained  since
simulation  tasks correspond to 2/3 of the Grid capacity.

\subsection{Interfacing PanDA with Titan}

The Titan supercomputer, current number two (number one from November 2012 until
June 2013) on the Top 500 list is located at the Oak Ridge Leadership Computing
Facility (OLCF) in Oak Ridge National Laboratory, US\@. Titan, was the first
large-scale system to use  a hybrid architecture that utilizes worker nodes with
both AMD 16-core Opteron 6274 CPUs and NVIDIA Tesla K20 GPU accelerators.

Integration with Titan is the current focus for PanDA developers. The project
aims to integrate Titan with the PanDA system using an updated PanDA Pilot that
runs on the front-end node and submits ATLAS payloads to the worker nodes using
the local batch system (PBS) via the SAGA (Simple API for Grid Applications)
interface~\cite{SAGA}. This solves several complications of  running on HPC
worker nodes,  including the lack of connectivity to outside world. The pilot
can communicate with the PanDA server from the front-end machine. The
interactive front-end  machines and the worker nodes  use  a shared  file system
which makes  it  easy  for the pilot to stage-in  any input files that are
required by the payload and stage-out the produced output files after the
payload has finished at the end of the job. ATLAS Tier 1 computing  center at
Brookhaven National Lab is currently  used for data stage out from Titan but in
principle  that can be any Grid site.

\paragraph*{HPC Backfill} HPC facilities are geared towards  large scale jobs by
design. Time allocation on an HPC is competitive and large projects are often
preferred. About 10\% of capacity on a  typical HPC machine  is unused  due to
mismatches between job sizes and available  resources. The worker nodes sit idle
because there are not enough of them or they do not have enough time to handle a
large scale computing job. On a machine of the scale of Titan, these 10\%
correspond to estimated 300M core hours per year. A system that can occupy those
temporarily free nodes with smaller scale tasks would be very valuable. This
offers a great possibility for PanDA to harvest the opportunistic resources on
Titan and use a similar mechanism on other HPCs. Functionality has been added to
the PanDA Pilot to collect information about available unused worker nodes on
Titan in real time. This allows PanDA to define precisely size and duration of
jobs submitted to Titan according to available free resources. First test of the
system were quite successful and show great promise in increasing the resource
utilization on Titan.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/PandaInterfaceWithHPC.png}
\caption{Schematic view of the PanDA system\label{fig:interface}}
\end{center}
\end{figure}

>>>>

The next LHC data taking run (Run 3) will require more resources than the
Worldwide LHC Computing Grid (WLCG) can possibly provide. Currently, PanDA WMS
uses more than 100,000 cores at over 100 Grid sites, with a peak performance of
0.3 petaFLOPS. This capacity will be sufficient for the planned analysis and
data processing, but it will be insufficient for Monte Carlo (MC) production and
any extra activities. To alleviate these challenges, ATLAS is engaged in a
program to expand the current computing model to include additional resources
such as the opportunistic use of supercomputers as well as commercial and
academic clouds.

% \subsection{Use of Supercomputers with PanDA}

Modern supercomputers have been designed to mainly support parallel computation
requiring runtime communication. Job execution is parallelized across multiple
cores, each core calculating a small part of the problem and communicating with
other cores via message passing interface (MPI). Accordingly, supercomputers
have large number of worker nodes, connected through a high-speed, low-latency
network. Each worker node has multicore CPUs, usually augmented with massively
parallel Graphics Processing Units (GPUs) or other types of coprocessors.

PanDA WMS has been designed to support distributed computing. Workload or
workflow executions involves concurrent and/or sequential runs of possibly large
amount of jobs, requiring no or minimal parallelization and no communication at
runtime. Accordingly, computing infrastructure like WLCG are designed to
aggregate large amount of computing resources across sites, possibly dislocated
across multiple geographical allocation. While each site may deploy runtime
message-passing capabilities, usually these are not used to perform distributed
computations.

There are at least two approaches to enable PanDA WMS to execute ATLAS jobs on
supercomputers: using the subset of resources and capabilities shared by both
supercomputers and grid; reconciling the two paradigms of computation by means
prototype a working solution that enables computation and better understanding
of the problem space. The latter approach is principled and better suited for a
production-grade solution developed on the base of the experience accumulated
with the prototype and capable of supporting general-purpose workloads and
workflows on both supercomputers and grid infrastructures.

This section illustrates the design, architecture and execution process of the
prototype developed by the PanDA team to support execution of some of the ATLAS
workload on the Titan supercomputer. After a critical analysis of the results
obtained and the lessons learned, the following section introduces the design
and first experimental characterization of a next generation executor. This
components is designed to abstract resources and capabilities enabling the
concurrent execution of both parallel and distributed computing.

% -----------------------------------------------------------------------------
\subsection{Interfacing PanDA with Titan}
\label{ssec:panda_titan}

The Titan supercomputer, current number two (number one from November 2012 until
June 2013) on the Top 500 list~\cite{top500} is located at the Oak Ridge
Leadership Computing Facility (OLCF) in Oak Ridge National Laboratory, US\@.
Titan is a Cray XK7 system with 18,688 compute nodes and a total of 299,008 CPU
cores. Each compute node has an AMD Opteron CPU, a Nvidia Tesla GPU, 32 GB of
RAM and no local storage, though a 16 GB RAM disk can be set up. Compute nodes
use Cray’s Gemini interconnect for inter-node MPI messaging but have no network
connection to the outside world. Titan is served by the Spider
II~\cite{spider2}, a Lustre filesystem with 32 PB of disk storage, and by a 29
PB HPSS tape storage system. Titan’s worker nodes run Compute Node Linux, a run
time environment based on SUSE Linux Enterprise Server.

% Titan, was the first large-scale system to use a hybrid architecture that
% utilizes worker nodes with both AMD 16-core Opteron 6274 CPUs and NVIDIA Tesla
% K20 GPU accelerators.
% This hybrid design provides improved energy efficiency, as well as an order of
% magnitude in computational capacity over its predecessor.

Titan offers to its users login nodes and data transfer nodes (DTN). Users have
to log into one of these nodes via ssh to submit jobs to Titan's PBS scheduler.
Titan's authentication and authorization model is based on two-factor
authentication with a RSA SecurID key, generated every 30 seconds. Login nodes
and DTN are the only nodes with out/inbound wide area network connectivity. Work
nodes have only local network access. Data staging is supported via the DTNs,
from the wide area network and the OLCF high performance storage system (HPSS).
DTNs are shared across all Titan's users that can schedule jobs to a dedicates
PBS batch system to automate data staging. Fair-share policies are in place both
for the batch system that for the use of DTN resources.

Titan's architecture, configuration and policies poses several challenges to the
integration with PanDA. PanDA Pilot requires to contact the Job Dispatcher
module of the PanDA Server to pull jobs to execute. Titan's work nodes do not
offer outbound network connectivity making the current deployment model of PanDA
pilots unfeasible. Titan does not support PanDA's security model based on
certificates and virtual organizations, making the PanDA's approach to identity
management also unfeasible. While DTNs offer wide area network data transfer, an
integration with ATLAS DDM is beyond the functional and administrative scope of
the current prototyping phase. Finally, the specific characteristics of the
execution environment, especially the dependences on a Lustre filesystem and on
modules tailored to Compute Node Linux, require reengineering of ATLAS
application frameworks.

Currently, no HEP application can benefit from Titan's GPUs but some
computationally-intensive, non memory-intensive tasks can be offloaded from the
grid to Titan's large amount of cores. Further, when HEP tasks can be
partitioned to independent jobs, Titan work nodes can be used to execute up to
16 concurrent jobs' payload. Given these constraints and challenges, the type of
task most suitable for execution on Titan is HEP event generation. Event
generators are mostly computational and stand-alone code, requiring less than
1GB of RAM at runtime and with small input data requirements. The event
generation in ATLAS accounts for 10--15\% of all jobs on the Grid, corresponding
to 2/3 of the current grid capacity. A more difficult workload to adapt for
Titan would be that for detector simulation. Simulation tasks tend to be tied to
the framework of a particular simulation, often requiring database access for
geometry and relevant run conditions.

\mtnote{Describe Monte Carlo workflow and Geant4 here.}

% While this is different from the HEP computing paradigm, where jobs are
% independent, it still shares common features such as the use of
% parallelization. It is not a requirement that HPC machines are able to run
% any possible task, nor is it relevant how many kinds of job types that can be
% run. What matters is the total number of cycles that can be offloaded from
% the Grid.

% Standard ATLAS workflow can not be easily ported to supercomputers due to
% several complications such as specialized worker node setups, no outbound
% network connections, limited memory per node, custom operating systems, etc. A
% reorganization of the standard workflow is therefore needed.

% -----------------------------------------------------------------------------
\subsection{PanDA Broker on Titan}
\label{ssec:panda_titan}

The most relevant design constraint for integrating PanDA WMS and Titan is the
lack of connectivity from the work nodes. This makes impossible to port PanDA
Pilot while maintaining the defining feature of the pilot abstraction:
multi-stage scheduling to decouple resource acquisition from workload scheduling
and execution. As direct communication between PanDA Pilot and PanDA Server is
required, PanDA Pilots cannot be scheduled on Titan and jobs cannot be pulled by
a pilot from the work nodes.

Usually, this design limitation would make the integration of PanDA and Titan
unfeasible. Pilot-based executions are necessary to avoid waiting in the
supercomputer's queue to execute every single job of a distributed workload.
This is particularly true for Titan where, consistently with the intended use
case of the machine, queuing policies privilege parallel jobs on the base of the
quantities of work nodes requested. It would still be possible to execute HEP
event generation tasks but the time spent waiting in the queue would be too
large compared to the execution time.

About 10\% of Titan's capacity is unused due to mismatches between job sizes and
available resources~\cite{titan_utilization}. Worker nodes are unused because
there are not enough nodes to execute large jobs or the required amount of nodes
would not be available for a long enough time. This makes available around
30,000 cores for a total of approximately 270M core hours per
year~\cite{titan_free}, roughly 30\% of the overall capacity of WLCG. A system
that can occupy those temporarily free nodes with smaller scale tasks would be
very valuable. For this reason, Titan offers backfill functionalities: users can
interrogate Titan's Moab scheduler about how many free nodes are currently
available and for how long. If a user submit a request for that number of nodes,
queue time will be minimal.

\mtnote{Please feel free to add details about backfill as needed.}

Backfill functionality offers the opportunity not to use the pilot abstraction
while still avoiding too much queue time overhead. On Titan, PanDA Pilot has
been redesigned to serve as a job broker. Architected as a stand-alone service
to be executed on a DTN, this PanDA Broker implements functionalities to: (i)
interrogate Titan about backfill availability; (ii) pull Geant4 jobs and events;
(iii) wrap Geant4 payloads into PBS MPI job; (iv) submitting PBS jobs to Titan's
batch system; and (v) staging input/output files. The new functionalities of
PanDA Broker involve backfill querying, payload wrapping, and job submission.
Functionalities for Geant4 job and events pulling, and file staging were mostly
inherited from PanDA Pilot.

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{figures/PanDA_setup_at_OLCF.png}
    \caption{Schematic view of the PanDA Broker.}
  \end{center}
\label{fig:panda_broker}
\end{figure}

Backfill querying is performed via the dedicated Moab command while payload
wrapping is done via a tailored Python MPI script. This MPI wrapper serves as a
``container'' for non-MPI jobs, enabling execution of unmodified Grid-centric
jobs on Titan. Typically, a MPI wrapper is workload-specific as it sets up the
execution environment for a specific payload. This involves organization of
worker directories, data management, optional input parameters modification, and
cleanup on exit. Upon submission, a copy of the MPI wrapper runs on every
available work node. Each script knows its MPI rank (an index that runs from
zero to a maximum number of nodes or script copies) as well as the total number
of ranks in the current submission. When activated on worker nodes, each copy of
the wrapper script starts the execution of the payload in a subprocess and waits
until its completion.

MPI wrappers are submitted to Titan's PBS batch system via
RADICAL-SAGA~\cite{radical-saga}. SAGA (Simple API for Grid Applications)
provides a unified interface to diverse job schedulers and file transferring
services. SAGA provides an interoperability layer that lowers the complexity and
improves the simplicity of using distributed infrastructure. RADICAL-SAGA is a
Python module, compliant with the OGF GFD.90 SAGA
specification~\cite{saga-spec}. Behind the API façade, RADICAL-SAGA implements a
adaptor architecture: each adaptor interface the SAGA API with different
middleware systems and services, including the batch scheduler of Titan.

PanDA Broker stages capabilities are implemented via a file system shared among
DTNs and work nods. Input files that are required by the payload are downloaded
via the ATLAS DDM service and then stored in a directory on the shared
filesystem. Part of the MPI wrapper setup process is to make the location of
these files known to the payload. Payloads' output files are located on the
shared filesystem by the broker and transferred from Titan to the ATLAS Tier 1
computing center at Brookhaven National Lab (BNL). In principle any Grid site
could be used.

The execution process of Geant4 on Titan by PanDA has X steps: (i) PanDA Broker
queries Titan's Moab scheduler about currently unused nodes and checks whether
the time and size of the available backfill resources are suitable to execute
jobs of Geant4 tasks; (ii) The PanDA Broker transmits this information to the
Job Dispatcher module of the PanDA server, and in response gets a list of Geant4
jobs bound to Titan; (iii) based on the jobs attributes, the broker requests to
the Data Service module of the PanDA Server to transfer the necessary input data
via the ATLAS DDM service; (iv) once all the data is transferred, the PanDA
Broker calculates the number of cores and duration of jobs that can be submitted
to Titan on the base of the information collected about backfill resource
availability; (v) the broker creates one or more PBS MPI scripts to wrap enough
Geant4 payload to fit the calculated number and duration of cores; (vi) the
broker submits jobs to Titan's PBS batch system via RADICAL-SAGA; (vii) the PBS
jobs waits a minimal amount of time into the queue until is scheduled for
execution; (viii) upon execution on the worker nodes, the MPI script sets up the
execution environment required by the wrapped Geant4 payload, including
assembling the AthenaMP framework in ramdisk; (ix) the payload is then executed
into dedicated processes, 16 for each work node, 1 for each core; (x) the PanDA
Broker locates on the shared filesystem the output files produced by the payload
and transfer them to BNL ATLAS Tier 1 computing centre.

% from the Job Dispatcher module of PanDA Server (as done by a PanDA Pilot);
% (iii) downloading events via PanDA Server's Data Service from ATLAS DDM to
% Titan's DTN; (iv) wrap the Geant4 payload into a PBS job; (v) submit jobs to
% Titan's PBS batch scheduler;

% Integration with Titan is the current focus for PanDA developers.

% The project aims to integrate Titan with the PanDA system using an updated
% PanDA Pilot that runs on the front-end node and submits ATLAS payloads to the
% worker nodes using the local batch system (PBS) via the SAGA (Simple API for
% Grid Applications) interface~\cite{SAGA}. This solves several complications
% of running on HPC worker nodes, including the lack of connectivity to outside
% world. The pilot can communicate with the PanDA server from the front-end
% machine.

% \paragraph*{HPC Backfill} HPC facilities are geared towards large scale jobs
% by design. Time allocation on an HPC is competitive and large projects are
% often preferred. About 10\% of capacity on a typical HPC machine is unused
% due to mismatches between job sizes and available resources. The worker nodes
% sit idle because there are not enough of them or they do not have enough time
% to handle a large scale computing job. On a machine of the scale of Titan,
% these 10\% correspond to estimated 300M core hours per year. A system that
% can occupy those temporarily free nodes with smaller scale tasks would be
% very valuable.

% This offers a great possibility for PanDA to harvest the opportunistic
% resources on Titan and use a similar mechanism on other HPCs. Functionality
% has been added to the PanDA Pilot to collect information about available
% unused worker nodes on Titan in real time.

% First test of the system were quite successful and show great promise in
% increasing the resource utilization on Titan.

### 2017-03-05 overall.tex
First pass IV (former IV and V) + restructuring

\subsection{PanDA Shared Library I/O Performance Impact at OLCF}

>>>>

% -----------------------------------------------------------------------------
\subsection{Geant4 Tasks Performance Analysis on Titan}
\label{ssec:panda_titan}

Currently, 20 instances of PanDA Broker are concurrently deployed on a total of
4 DTNs, 5 instances for each DTN. Each broker controls from 15 to 300 ATLAS
simulation jobs per submission, for a maximum total of 96,000 cores. Since
November 2015, the brokers have operated only in backfill mode, without a
defined time allocation, and running at the lowest priority on Titan.
Figure~\ref{core-hours-utilization} shows Titan core hours used by ATLAS from
September 2015 to March 2017. During that period, ATLAS consumed about XX
million Titan core hours. About XX million detector simulation jobs were
completed, and more than 132 million events processed. In February 2017, we
exceeded X million core hours per month and by March 2017 we ran at more than
X million core hours per month.

Figure~\ref{backfill-utilization} shows backfill utilization efficiency, defined
as a fraction of Titan’s total free resources that were utilized by ATLAS, for
the period from September 2015 to March 2017. Due to continuos improvements in
the system, the efficiency grew from less than 5\% to more than XX\%, reaching
XX\% in February 2017. At the peak, this corresponds to more than X\%
improvement in Titan’s overall utilization.

\mtnote{Add new diagram as discussed in F2F meeting at Rutgers.}

% -----------------------------------------------------------------------------
\subsection{PanDA Shared Library I/O Performance Impact at OLCF}

### 2017-03-05 titan.tex
First pass IV (former IV and V) + restructuring

In 2004, the Oak Ridge Leadership Computing Facility (OLCF) was established at
Oak Ridge National Laboratory (ORNL).The  OLCF's is mission to accelerate the
scientific discovery and engineering progress by providing outstanding
computing and data management resources to high-priority research and
development projects.

Titan, a leadership-class and the fastest open-science supercomputer in the
U.S., is located at OLCF. Titan enables scientists to evaluate and assess various
complex physical phenomena via large-scale computational simulations.

Titan is a Cray XK7 system with 18,688 compute nodes~\cite{top500}. Each
compute node is equipped with an AMD Opteron CPU and a Nvidia Tesla GPU. This
hybrid design provides improved energy efficiency, as well as an order of
magnitude in computational capacity over its predecessor. Titan has 710 TB of
total system memory. This leadership capability compute system is backed by a
center-wide parallel file system known as Spider II~\cite{spider2}. Spider II
is one of the world's fastest and largest POSIX complaint parallel file
systems. It is designed to serve write-heavy I/O workloads generated by Titan
compute clients and other OLCF resources.

>>>>

In 2004, the Oak Ridge Leadership Computing Facility (OLCF) was established at
Oak Ridge National Laboratory (ORNL). The OLCF's mission is to accelerate
scientific discovery and engineering progress by providing outstanding computing
and data management resources to high-priority research and development
projects. Among other resources, OLCF manages Titan, a leadership-class and the
fastest open-science supercomputer in the US. Titan enables scientists to
evaluate and assess various complex physical phenomena via large-scale
computational simulations.

Titan is a Cray XK7 system with 18,688 compute nodes~\cite{top500}. Each compute
node is equipped with an AMD Opteron CPU and a Nvidia Tesla GPU. This hybrid
design provides improved energy efficiency, as well as an order of magnitude in
computational capacity over its predecessor. Titan has 710 TB of total system
memory and a center-wide parallel file system known as Spider II~\cite{spider2}.
Spider II is one of the world's fastest and largest POSIX complaint parallel
file systems, designed to serve write-heavy I/O workloads generated by Titan
compute clients and other OLCF resources.
